---
title: "Chd1 RNAseq GATK variant"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mutational signatures analysis of Chd1 RNAseq 

This is using the RNAseq data from CHD1 WT and Chd1-R616Q E18.5 NPCs from our mouse model.

### First, we aligned the RNAseq data to the genome using STAR 2 pass

```{bash, eval = FALSE}
#!/bin/sh

#SBATCH --mail-type ALL
#SBATCH --mail-user=kimberl@hi.is
#SBATCH --job-name=STAR     # Job name
#SBATCH --output=/hpcdata/Mimir/kimberl/logfiles/20241025/20241025_STAR2pass.out    # Output file (STDOUT)
#SBATCH --error=/hpcdata/Mimir/kimberl/logfiles/20241025/202410025_STAR2pass.err   # Error file (STDERR)
#SBATCH --partition=128cpu_256mem
#SBATCH -N 1
#SBATCH -n 128

## ran with this command:  sbatch ...


cd /proj/hpcdata/Mimir/kimberl/20230102_Novogene_RNAseq/CHD1_RNAseq

find . -maxdepth 1 -mindepth 1 -type d -exec bash -c '
    cd "$1" && pwd &&
    name=$(basename "$PWD")
    
    # Check if there are *_1.fq.gz and *_2.fq.gz files
    if ls *_1.fq.gz >/dev/null 2>&1 && ls *_2.fq.gz >/dev/null 2>&1; then
        # Get comma-separated list of *_1.fq.gz files
        r1_files=$(ls *_1.fq.gz | paste -sd "," -)
        
        # Get comma-separated list of *_2.fq.gz files
        r2_files=$(ls *_2.fq.gz | paste -sd "," -)
        
        # Run STAR command with comma-separated lists of files
        STAR --runThreadN 128 \
        --genomeDir /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/STAR_index \
        --sjdbGTFfile /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.112.chr.gtf \
        --readFilesIn "$r1_files" "$r2_files" \
        --readFilesCommand zcat \
        --outWigType wiggle \
        --twopassMode Basic \
        --outFileNamePrefix 20241024_STAR_2pass/"${name}_STAR" \
        --outSAMtype BAM SortedByCoordinate
    else
        echo "Skipping $name: Missing *_1.fq.gz or *_2.fq.gz files."
    fi
' bash {} \;

```

### Running GATK

```{bash, eval = FALSE}
#!/bin/sh

#SBATCH --job-name=gatk     # Job name
#SBATCH --mail-type ALL
#SBATCH --mail-user=kimberl@hi.is
#SBATCH --output=/hpcdata/Mimir/kimberl/logfiles/20241029/20241029_GATK_h.out    # Output file (STDOUT)
#SBATCH --error=/hpcdata/Mimir/kimberl/logfiles/20241029/20241029_GATK_h.err   # Error file (STDERR)
#SBATCH --partition=48cpu_192mem
#SBATCH -N 3
#SBATCH -n 144

ml use /hpcapps/libsci-gcc/modules/all
ml use /hpcapps/lib-mimir/modules/all
ml use /hpcapps/libbio-gpu/modules/all


ml load Java/15.0.1 
ml load picard

# Loop over each .bam file in the folder
for input_file in *.sortedByCoord.out.bam; do
    # Define output file by appending "_filtered" to the original filename
    output_file="${input_file%.sortedByCoord.out.bam}_out"

    java -jar $EBROOTPICARD/picard.jar MarkDuplicates \
        I="$input_file" \
        O="$output_file"_marked_duplicates.bam \
        M="$output_file"_marked_dup_metrics.txt \
    
    echo "Processed $input_file"
done



ml load  Java/17.0.6
module load GATK/4.4.0.0-GCCcore-12.2.0-Java-17
ml load SAMtools/1.17-GCC-11.2.0

for input_file in *STARAligned_out_marked_duplicates.bam; do
    # Define output file by appending "_filtered" to the original filename
    output_file="${input_file%_STARAligned_out_marked_duplicates.bam}_out"

    gatk SplitNCigarReads \
        -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
        -I "$input_file" \
        -O "$output_file"_SplitNCigar_output.bam
    done
        
for input_file in *_out_SplitNCigar_output.bam; do
    # Define output file by appending "_filtered" to the original filename
    output_file="${input_file%_out_SplitNCigar_output.bam}"
    
    samtools addreplacerg -r "ID:$output_file" -r 'LB:library1' -r 'PL:ILLUMINA' -r 'PU:unit1' -r "SM:$output_file" \
    -o RG_"$output_file".bam "$input_file" -@ 144
    samtools index RG_"$output_file".bam -@ 96

    gatk BaseRecalibrator \
        -I RG_"$output_file".bam \
        -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
        --known-sites /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/mus_musculus.vcf \
        --use-original-qualities \
        -O "$output_file"_recal_data.table

     gatk ApplyBQSR \
        -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
        -I RG_"$output_file".bam \
        --bqsr-recal-file "$output_file"_recal_data.table \
        --add-output-sam-program-record \
        --use-original-qualities \
        -O "$output_file"_bsqr_output.bam

    gatk AnalyzeCovariates \
        -bqsr "$output_file"_recal_data.table \
        -plots "$output_file"_AnalyzeCovariates.pdf

    
    done

# Step 1. Run Mutect2 in tumor-only mode for each normal sample.

gatk Mutect2 -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
    -I A1470_3_DMSO_bsqr_output.bam \
    -max-mnp-distance 0 \
    --native-pair-hmm-threads 144 \
    -O normal1.vcf.gz

gatk Mutect2 -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
    -I A1470_4_DMSO_bsqr_output.bam \
    -max-mnp-distance 0 \
    --native-pair-hmm-threads 144 \
    -O normal2.vcf.gz

gatk Mutect2 -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
    -I A2288_6_DMSO_bsqr_output.bam \
    -max-mnp-distance 0 \
    --native-pair-hmm-threads 144 \
    -O normal3.vcf.gz

gatk Mutect2 -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
    -I A2288_8_DMSO_bsqr_output.bam \
    -max-mnp-distance 0 \
    --native-pair-hmm-threads 144 \
    -O normal4.vcf.gz

# Step 2. Create a GenomicsDB from the normal Mutect2 calls.

gatk GenomicsDBImport -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
        --genomicsdb-workspace-path pon_db \
        -L /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39_intervals.list \
        -V normal1.vcf.gz \
        -V normal2.vcf.gz \
        -V normal3.vcf.gz \
        -V normal4.vcf.gz

# Create panel of normals.

 gatk CreateSomaticPanelOfNormals -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
     -V gendb://pon_db -O pon.vcf.gz


# Now call variants using the panel of normals and the germline resource
for input_file in *_bsqr_output.bam; do
    
    file_name=$(basename "$input_file")
    name="${file_name%_bsqr_output.bam}"
    
    gatk Mutect2 \
    -R /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/Mus_musculus.GRCm39.dna.primary_assembly.fa \
    -I "$input_file" \
    --germline-resource /proj/hpcdata/Mimir/kimberl/Genome_indexes/GRCm39_v112/mus_musculus.vcf \
    --panel-of-normals pon.vcf.gz \
    --native-pair-hmm-threads 144 \
    -O "$name"_vs_PoN.vcf.gz

done
```

### Filtering vcf files using bcftools isec to remove variants that are present in at least 12 of the 16 samples
```{bash, eval=FALSE}

#!/bin/sh

#SBATCH --job-name=gatk     # Job name
#SBATCH --mail-type ALL
#SBATCH --mail-user=kimberl@hi.is
#SBATCH --output=/hpcdata/Mimir/kimberl/logfiles/20241105/20241105.out   # Output file (STDOUT)
#SBATCH --error=/hpcdata/Mimir/kimberl/logfiles/20241105/20241105.err  # Error file (STDERR)
#SBATCH --partition=48cpu_192mem
#SBATCH -N 2
#SBATCH -n 96

ml use /hpcapps/libsci-gcc/modules/all
ml use /hpcapps/lib-mimir/modules/all
ml use /hpcapps/libbio-gpu/modules/all

ml load  BCFtools/1.17-GCC-11.2.0

# Start by merging all vcf files to find the common variants. This will be made in the site.vcf.txt file
bcftools isec -p 20241105_filtered_output/ -n +12 -c all -Oz *.vcf.gz


# Now Filter each VCF file in a loop
for vcf in *.vcf.gz; do
    bcftools view -T ^20241105_filtered_output/sites.txt $vcf -Oz -o filtered_${vcf}
    bcftools index filtered_${vcf}
done
```

## RNAseq variant analysis after GATK

```{r variants, warning=FALSE, cache=TRUE}
library(tidyverse)
library(MutationalPatterns)
library(BSgenome.Mmusculus.UCSC.mm39)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm39.refGene)
library(NMF)

bsGenome <- BSgenome.Mmusculus.UCSC.mm39
txdb <- TxDb.Mmusculus.UCSC.mm39.refGene

# Read filtered VCF files
pattern <- "\\.vcf\\.gz$|\\.vcf$"

vcf_files <- list.files(
  path = "STAR_2pass_vcf/Filtered/",
  pattern = pattern,
  full.names = TRUE
)

filenames <- basename(vcf_files)
filenames <- sub("\\..*$", "", filenames)

# Read indels and SNVs separately
indels <- read_vcfs_as_granges(
  vcf_files,
  sample_names = filenames,
  genome = "BSgenome.Mmusculus.UCSC.mm39",
  type = "indel"
)

snvs <- read_vcfs_as_granges(
  vcf_files,
  sample_names = filenames,
  genome = "BSgenome.Mmusculus.UCSC.mm39",
  type = "snv"
)

# Limit to exonic variants
exons <- exons(txdb)

indels_exonic <- lapply(indels, function(gr) subsetByOverlaps(gr, exons))
snvs_exonic   <- lapply(snvs,   function(gr) subsetByOverlaps(gr, exons))

# Count exonic indels and SNVs per sample
indelnum <- sapply(indels_exonic, length)
snvnum   <- sapply(snvs_exonic, length)

write.csv2(data.frame(indelnum), "indelnum.csv")
write.csv2(data.frame(snvnum), "SNVnum.csv")

# Convert indels to GRangesList for MutationalPatterns
indels_exonic <- GRangesList(indels_exonic)

# Get indel contexts and counts
indel_context <- get_indel_context(
  indels_exonic,
  ref_genome = "BSgenome.Mmusculus.UCSC.mm39"
)

indel_counts <- count_indel_contexts(indel_context)

write.csv2(indel_counts, "indel_context.csv")

# Fit exonic indel counts to known COSMIC indel signatures
indel_counts1 <- indel_counts + 0.0001

signatures <- get_known_signatures("indel", source = "COSMIC")

fit_res <- fit_to_signatures(indel_counts1, signatures)

write.csv(fit_res$contribution, "COSMIC_sig.csv")
```

## COSMIC indel signature statistical analysis

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)

df_cosmic_all <- read.csv(
  "COSMIC_sig.csv",
  header = TRUE,
  check.names = FALSE,
  row.names = 1,
  fileEncoding = "latin1"
)

genotype_map <- c(
  "A1470_1" = "Chd1+/R616Q",
  "A1470_3" = "Chd1+/+",
  "A1470_4" = "Chd1+/+",
  "A1470_5" = "Chd1+/R616Q",
  "A2288_2" = "Chd1+/R616Q",
  "A2288_5" = "Chd1+/R616Q",
  "A2288_6" = "Chd1+/+",
  "A2288_8" = "Chd1+/+"
)

df_cosmic_long <- df_cosmic_all %>%
  rownames_to_column("Signature") %>%
  pivot_longer(-Signature, names_to = "Sample", values_to = "Count") %>%
  mutate(
    sample_key = str_extract(Sample, "A\\d+_\\d+"),
    Genotype = genotype_map[sample_key],
    Treatment = ifelse(grepl("DHT", Sample), "DHT", "DMSO"),
    Genotype = factor(Genotype, levels = c("Chd1+/+", "Chd1+/R616Q")),
    Treatment = factor(Treatment, levels = c("DMSO", "DHT"))
  )

df_dmso <- df_cosmic_long %>%
  filter(Treatment == "DMSO")

signatures_all <- unique(df_cosmic_long$Signature)

results_all <- do.call(rbind, lapply(signatures_all, function(sig) {
  wt <- df_dmso$Count[df_dmso$Signature == sig & df_dmso$Genotype == "Chd1+/+"]
  mut <- df_dmso$Count[df_dmso$Signature == sig & df_dmso$Genotype == "Chd1+/R616Q"]
  
  if (sd(c(wt, mut)) == 0 || length(wt) < 2 || length(mut) < 2) {
    return(data.frame(
      Signature = sig,
      mean_WT = mean(wt),
      mean_Mut = mean(mut),
      t = NA,
      df = NA,
      p_raw = NA,
      eta2 = NA
    ))
  }
  
  tt <- t.test(wt, mut, var.equal = FALSE)
  eta2 <- tt$statistic^2 / (tt$statistic^2 + tt$parameter)
  
  data.frame(
    Signature = sig,
    mean_WT = round(mean(wt), 3),
    mean_Mut = round(mean(mut), 3),
    t = round(tt$statistic, 4),
    df = round(tt$parameter, 3),
    p_raw = round(tt$p.value, 6),
    eta2 = round(eta2, 4)
  )
}))

results_all <- results_all %>%
  mutate(
    p_holm = round(p.adjust(p_raw, method = "holm"), 6),
    significant = p_holm < 0.05
  ) %>%
  arrange(p_raw)

results_all %>%
  mutate(
    p_raw = signif(p_raw, 3),
    p_holm = signif(p_holm, 3),
    significant = ifelse(significant, "Yes", "No")
  ) %>%
  kable(
    caption = "COSMIC indel signature comparison between Chd1+/+ and Chd1R616Q/+ DMSO samples",
    digits = 3
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(which(results_all$significant), bold = TRUE)
```




