---
title: "Chd1^R616Q^ Mouse — Statistical Analysis"
author: "Kimberley Jade Anderson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi = 150
)
```

```{css, echo=FALSE}
/* ── Typography ── */
body   { font-family: "Source Sans Pro", sans-serif; font-size: 15px; }
h1, h2 { color: #2c3e50; }
h3     { color: #5960AB; }

/* ── Callout boxes ── */
.note {
  background: #eef3fb; border-left: 4px solid #5960AB;
  padding: 10px 15px; border-radius: 4px; margin: 12px 0;
}
.warn {
  background: #fff4e5; border-left: 4px solid #EE4142;
  padding: 10px 15px; border-radius: 4px; margin: 12px 0;
}

/* ── Code blocks ── */
pre { background: #f8f9fa; border-radius: 6px; }
```

# Overview

This document presents the statistical analyses underpinning the *Chd1^R616Q^* manuscript, covering:

- **Body-weight trajectories** — early postnatal, adult, and castrated cohorts
- **Effect-size summary** — partial η² and Cohen's *d* for Welch's *t*-tests
- **NPC cell-cycle composition** — MANOVA + univariate follow-ups
- **Embryonic testosterone treatment** — Ki67 proliferation by genotype × treatment

---

# Packages

```{r libraries}
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest)
```

---

# Helper Functions {.tabset}

## Timepoint ordering

```{r helper-order-timepoints}
order_timepoints <- function(tp) {
  num <- suppressWarnings(as.numeric(stringr::str_extract(tp, "\\d+")))
  if (all(is.na(num))) sort(unique(tp)) else unique(tp)[order(num, tp)]
}
```

## Partial η²

```{r helper-eta-sq}
partial_eta_sq <- function(anova_table) {
  results <- data.frame(
    Effect       = rownames(anova_table),
    F_value      = round(anova_table[["F value"]], 4),
    NumDF        = anova_table[["NumDF"]],
    DenDF        = round(anova_table[["DenDF"]], 3),
    p_value      = round(anova_table[["Pr(>F)"]], 6),
    partial_eta2 = NA_real_
  )

  ms_residual          <- anova_table[["Mean Sq"]] / anova_table[["F value"]]
  ss_residual          <- ms_residual * anova_table[["DenDF"]]
  ss_effect            <- anova_table[["Sum Sq"]]
  results$partial_eta2 <- round(ss_effect / (ss_effect + ss_residual), 4)

  results$interpretation <- cut(
    results$partial_eta2,
    breaks = c(-Inf, 0.01, 0.06, 0.14, Inf),
    labels = c("negligible", "small", "medium", "large")
  )
  results
}
```

## Main analysis function

```{r helper-main-fn}
read_weights_and_analyze <- function(
    csv_file,
    sep            = ",",
    id_col         = NULL,
    sex_col        = "Sex",
    genotype_col   = "Genotype",
    output_prefix  = "weights",
    plot_width     = 6.2,
    plot_height    = 4.2,
    crossbar_width = 2,
    breaks         = breaks,
    point_dodge    = 2,
    point_jitter   = 0.8,
    pos            = position_identity()) {

  # 1) Read
  df <- read.delim(csv_file, sep = sep, header = TRUE, check.names = FALSE)

  required    <- c(sex_col, genotype_col)
  missing_req <- setdiff(required, names(df))
  if (length(missing_req) > 0)
    stop("Missing required column(s): ", paste(missing_req, collapse = ", "))

  # 2) Ensure MouseID exists
  if (!is.null(id_col) && id_col %in% names(df)) {
    df <- df %>% rename(MouseID = all_of(id_col))
  } else if (!("MouseID" %in% names(df))) {
    df <- df %>% mutate(MouseID = paste0(.data[[sex_col]], "_", row_number()))
  }

  # 3) Detect timepoint columns
  meta_cols <- c(sex_col, genotype_col, "MouseID")
  time_cols <- setdiff(names(df), meta_cols)
  if (length(time_cols) < 2)
    stop("Need at least 2 timepoint columns. Detected: ", paste(time_cols, collapse = ", "))

  # 4) Long format
  df_long <- df %>%
    pivot_longer(cols = all_of(time_cols), names_to = "Timepoint", values_to = "Weight") %>%
    mutate(
      Sex      = as.factor(.data[[sex_col]]),
      Genotype = as.factor(.data[[genotype_col]]),
      Weight   = as.numeric(Weight),
      Age      = as.numeric(stringr::str_extract(Timepoint, "\\d+"))
    )

  # 5) Order timepoints
  tp_levels <- order_timepoints(df_long$Timepoint)
  df_long   <- df_long %>% mutate(Timepoint = factor(Timepoint, levels = tp_levels))

  # 6) Report dropped rows
  dropped_rows <- df_long %>% filter(is.na(Weight) | is.na(Age))
  if (nrow(dropped_rows) > 0) {
    message("\n⚠️  Rows excluded (NA Weight or Age):")
    dropped_rows %>% count(Sex, Genotype, Timepoint, name = "n_missing") %>% print()
  } else {
    message("✓ No rows dropped.")
  }

  # 7) Summary stats
  df_sum <- df_long %>%
    group_by(Sex, Genotype, Timepoint, Age) %>%
    summarise(
      mean = mean(Weight, na.rm = TRUE),
      sd   = sd(Weight,  na.rm = TRUE),
      n    = sum(!is.na(Weight)),
      .groups = "drop"
    )

  # ── Plotting ──────────────────────────────────────────────────────────────
  plot_one_sex <- function(sex_letter) {
    p_dat        <- df_long %>% filter(Sex == sex_letter)
    p_sum        <- df_sum  %>% filter(Sex == sex_letter)
    geno_levels  <- levels(p_dat$Genotype)
    geno_offsets <- setNames(c(-point_dodge / 2, point_dodge / 2), geno_levels)

    ggplot() +
      geom_point(
        data = p_dat,
        aes(x = Age + geno_offsets[as.character(Genotype)],
            y = Weight, color = Genotype, shape = Genotype),
        position = pos, alpha = 0.87, size = 3, stroke = 1
      ) +
      geom_errorbar(
        data = p_sum,
        aes(x    = Age + geno_offsets[as.character(Genotype)],
            ymin = mean - sd, ymax = mean + sd, color = Genotype),
        width = 0.4, linewidth = 0.3
      ) +
      geom_line(
        data = p_sum,
        aes(x = Age + geno_offsets[as.character(Genotype)],
            y = mean, color = Genotype, group = Genotype),
        linewidth = 0.8
      ) +
      geom_crossbar(
        data = p_sum,
        aes(x = Age + geno_offsets[as.character(Genotype)],
            y = mean, ymin = mean, ymax = mean, group = Genotype),
        width = crossbar_width, linewidth = 0.45, color = "black"
      ) +
      scale_x_continuous(
        breaks = sort(unique(p_dat$Age)),
        labels = sort(unique(p_dat$Age))
      ) +
      scale_y_continuous(
        limits = c(0, NA),
        expand = expansion(mult = c(0, 0.05)),
        breaks = breaks
      ) +
      scale_color_manual(values = c(
        "Chd1+/+"     = "#5960AB",
        "Chd1+/R616Q" = "#EE4142"
      )) +
      scale_shape_manual(values = c(
        "Chd1+/+"     = 1,
        "Chd1+/R616Q" = 17
      )) +
      labs(
        title  = paste(sex_letter, "— weight over time"),
        x      = "Age (days)",
        y      = "Weight (g)",
        color  = "Genotype",
        shape  = "Genotype"
      ) +
      theme_classic(base_size = 20) +
      theme(
        axis.title        = element_text(size = 24, color = "black"),
        axis.text         = element_text(size = 20, color = "black"),
        axis.line         = element_line(linewidth = 0.8, color = "black"),
        axis.ticks        = element_line(linewidth = 0.8, color = "black"),
        axis.ticks.length = unit(0.22, "cm"),
        legend.position   = "right",
        legend.title      = element_text(size = 18),
        legend.text       = element_text(size = 16),
        plot.margin       = margin(10, 15, 10, 10)
      )
  }

  sexes <- sort(unique(as.character(df_long$Sex)))
  plots <- setNames(lapply(sexes, plot_one_sex), sexes)

  for (sx in names(plots)) {
    ggsave(
      filename = paste0(output_prefix, "_", sx, ".pdf"),
      plot     = plots[[sx]],
      width    = plot_width,
      height   = plot_height,
      dpi      = 300
    )
  }

  # ── Statistics ────────────────────────────────────────────────────────────
  model_all    <- lmer(Weight ~ Genotype * Timepoint + (1 | MouseID), data = df_long)
  model_by_sex <- lapply(sexes, function(sx)
    lmer(Weight ~ Genotype * Timepoint + (1 | MouseID),
         data = df_long %>% filter(Sex == sx)))
  names(model_by_sex) <- sexes

  list(
    df           = df,
    df_long      = df_long,
    df_sum       = df_sum,
    dropped_rows = dropped_rows,
    plots        = plots,
    model_all    = model_all,
    anova_all    = anova(model_all),
    model_by_sex = model_by_sex,
    anova_by_sex = lapply(model_by_sex, anova),
    eta_by_sex   = lapply(model_by_sex, function(m) partial_eta_sq(anova(m)))
  )
}
```

---

# Body-Weight Trajectories

## Early postnatal cohort

```{r young-mice, cache=TRUE}
young_mice <- read_weights_and_analyze(
  "Mouse_early_weights.csv",
  sep            = ";",
  output_prefix  = "young",
  plot_width     = 9,
  plot_height    = 5,
  crossbar_width = 0.3,
  point_dodge    = 0.4,
  point_jitter   = 0.001,
  pos            = position_identity(),
  breaks         = function(x) seq(0, x[2], by = 2)
)
```

### Plots {.tabset}

#### Female
```{r young-plot-F, fig.width=9, fig.height=5}
young_mice$plots[["F"]]
```

#### Male
```{r young-plot-M, fig.width=9, fig.height=5}
young_mice$plots[["M"]]
```

### Dropped observations
```{r young-dropped}
young_mice$dropped_rows
```

### Mixed-model ANOVA — effect sizes by sex
```{r young-eta}
young_mice$eta_by_sex
```

---

## Adult cohort

```{r adult-mice, cache=TRUE}
adult_mice <- read_weights_and_analyze(
  "adult_mice_weights.csv",
  sep            = ";",
  output_prefix  = "adult",
  plot_width     = 9,
  plot_height    = 5,
  crossbar_width = 3,
  point_dodge    = 3.8,
  pos            = position_jitter(width = 1, height = 0),
  breaks         = function(x) seq(0, x[2], by = 5)
)
```

### Plots {.tabset}

#### Female
```{r adult-plot-F, fig.width=9, fig.height=5}
adult_mice$plots[["F"]]
```

#### Male
```{r adult-plot-M, fig.width=9, fig.height=5}
adult_mice$plots[["M"]]
```

### Dropped observations
```{r adult-dropped}
adult_mice$dropped_rows
```

### Overall mixed-model ANOVA
```{r adult-anova-all}
adult_mice$anova_all
```

### Mixed-model ANOVA by sex
```{r adult-anova-sex}
adult_mice$anova_by_sex
```

### Effect sizes by sex
```{r adult-eta}
adult_mice$eta_by_sex
```

---

## Castrated cohort

```{r castrated-mice, cache=TRUE}
castrated <- read_weights_and_analyze(
  "castrated_weights_clean.csv",
  sep            = ";",
  output_prefix  = "castrated",
  plot_width     = 9,
  plot_height    = 5,
  crossbar_width = 3,
  point_dodge    = 1.5,
  pos            = position_jitter(width = 0.05, height = 0),
  breaks         = function(x) seq(0, x[2], by = 5)
)
```

### Plots {.tabset}

#### Female
```{r castrated-plot-F, fig.width=9, fig.height=5}
castrated$plots[["F"]]
```

#### Male
```{r castrated-plot-M, fig.width=9, fig.height=5}
castrated$plots[["M"]]
```

### Mixed-model ANOVA by sex
```{r castrated-anova}
castrated$anova_by_sex
```

### Effect sizes by sex
```{r castrated-eta}
castrated$eta_by_sex
```

---

# Effect-Size Summary (Welch's *t*-tests)

<div class="note">
η² values from individual Welch's *t*-tests are converted to Cohen's *d* via  
*d* = 2√(η² / (1 − η²)).  Benchmarks: negligible < 0.2, small 0.2–0.5, medium 0.5–0.8, large > 0.8.
</div>

```{r cohens-d-fn}
eta2_to_cohens_d <- function(eta2, label = NULL) {
  d <- 2 * sqrt(eta2 / (1 - eta2))
  interpretation <- cut(
    abs(d),
    breaks = c(-Inf, 0.2, 0.5, 0.8, Inf),
    labels = c("negligible", "small", "medium", "large")
  )
  result <- data.frame(
    eta2           = round(eta2, 4),
    cohens_d       = round(d,   4),
    interpretation = interpretation
  )
  if (!is.null(label)) result <- cbind(comparison = label, result)
  result
}
```

```{r welch-effects}
welch_effect_sizes <- rbind(
  eta2_to_cohens_d(0.3151, "Female surface righting"),
  eta2_to_cohens_d(0.2366, "Phenotype score: females vs males"),
  eta2_to_cohens_d(0.4558, "MWM velocity"),
  eta2_to_cohens_d(0.3744, "MWM distance"),
  eta2_to_cohens_d(0.7785, "Western blot protein"),
  eta2_to_cohens_d(0.2443, "Open field (female)"),
  eta2_to_cohens_d(0.5494, "MTT assay — female NPCs"),
  eta2_to_cohens_d(0.2534, "Oxytocin ELISA"),
  eta2_to_cohens_d(0.8667, "Indels in RNA")
)

welch_effect_sizes
```

---

# NPC Cell-Cycle Composition

## Data

```{r cell-cycle-data}
cell_cycle <- read.csv("NPC_Cell_cycle.csv",
                       row.names = 1, header = TRUE, check.names = FALSE)

genotype_labels <- colnames(cell_cycle)

df_cc <- as.data.frame(t(cell_cycle))
colnames(df_cc) <- c("G0G1", "S", "G2M")
df_cc$Genotype  <- factor(genotype_labels,
                           levels = c("Chd1+/+", "Chd1R616Q/+"))
rownames(df_cc)  <- NULL
df_cc <- df_cc %>% mutate(across(c(G0G1, S, G2M), as.numeric))

df_cc
```

## MANOVA

```{r cell-cycle-manova}
manova_fit     <- manova(cbind(G0G1, S, G2M) ~ Genotype, data = df_cc)
manova_summary <- summary(manova_fit, test = "Pillai")
manova_summary
```

```{r pillai}
pillai <- manova_summary$stats["Genotype", "Pillai"]
cat("Pillai's trace (overall effect size):", round(pillai, 4), "\n")
```

## Univariate follow-ups

```{r cell-cycle-univariate}
univariate_summary <- summary.aov(manova_fit)
univariate_summary
```

## Effect sizes per phase

```{r cell-cycle-eta}
cat("Available names in univariate_summary:\n")
print(names(univariate_summary))

phases <- c("G0G1", "S", "G2M")

uni_results <- do.call(rbind, lapply(phases, function(ph) {
  
  # Find the matching name flexibly
  key <- names(univariate_summary)[grepl(ph, names(univariate_summary))]
  
  if (length(key) == 0) {
    warning("No match found for phase: ", ph)
    return(NULL)
  }
  
  aov_table   <- univariate_summary[[key]]
  ss_effect   <- as.numeric(aov_table["Genotype",  "Sum Sq"])
  ss_residual <- as.numeric(aov_table["Residuals", "Sum Sq"])
  f_val       <- as.numeric(aov_table["Genotype",  "F value"])
  df_num      <- as.numeric(aov_table["Genotype",  "Df"])
  df_den      <- as.numeric(aov_table["Residuals", "Df"])
  p_val       <- as.numeric(aov_table["Genotype",  "Pr(>F)"])
  eta2        <- ss_effect / (ss_effect + ss_residual)
  d           <- 2 * sqrt(eta2 / (1 - eta2))

  data.frame(
    Phase    = ph,
    mean_WT  = round(mean(df_cc[df_cc$Genotype == "Chd1+/+",     ph], na.rm = TRUE), 3),
    mean_Mut = round(mean(df_cc[df_cc$Genotype == "Chd1R616Q/+", ph], na.rm = TRUE), 3),
    F_value  = round(f_val, 4),
    df_num   = df_num,
    df_den   = df_den,
    p_raw    = round(p_val, 6),
    eta2     = round(eta2,  4),
    cohens_d = round(d,     4)
  )
}))

uni_results
```


---

# Embryonic Testosterone Treatment — Ki67 FACS

## Data

```{r embryo-data}
df_raw <- read.csv("Embryo_T_Treatment.csv",
                   header = TRUE, check.names = FALSE, row.names = 1,
                   fileEncoding = "latin1")

genotype_labels  <- colnames(df_raw)
colnames(df_raw) <- make.unique(colnames(df_raw))
rownames(df_raw) <- c("Corn oil", "Testosterone")

df_ki67 <- df_raw %>%
  mutate(Treatment = trimws(rownames(.))) %>%
  pivot_longer(cols = -Treatment, names_to = "Sample", values_to = "Ki67_pct") %>%
  mutate(Genotype = rep(genotype_labels, times = nrow(df_raw))) %>%
  filter(!is.na(Ki67_pct) & Ki67_pct != "") %>%
  mutate(
    Ki67_pct  = as.numeric(Ki67_pct),
    Genotype  = factor(Genotype,  levels = c("Chd1+/+", "Chd1R616Q/+")),
    Treatment = factor(Treatment, levels = c("Corn oil", "Testosterone"))
  ) %>%
  select(Genotype, Treatment, Ki67_pct)

df_ki67

cat("\nSample sizes per group:\n")
print(table(df_ki67$Genotype, df_ki67$Treatment))
```

## Two-way ANOVA

```{r embryo-anova}
aov_fit     <- aov(Ki67_pct ~ Genotype * Treatment, data = df_ki67)
aov_summary <- summary(aov_fit)
aov_summary
```

## Effect sizes (η²)

```{r embryo-eta}
ss_table <- as.data.frame(aov_summary[[1]])
ss_total <- sum(ss_table[["Sum Sq"]], na.rm = TRUE)

effect_sizes <- data.frame(
  Effect  = trimws(rownames(ss_table)[1:3]),
  SS      = round(ss_table[["Sum Sq"]][1:3], 4),
  F_value = round(ss_table[["F value"]][1:3], 4),
  p_value = round(ss_table[["Pr(>F)"]][1:3], 6),
  eta2    = round(ss_table[["Sum Sq"]][1:3] / ss_total, 6)
)

effect_sizes
```

## Post-hoc comparisons (Tukey HSD)

```{r embryo-tukey}
TukeyHSD(aov_fit)
```

## Post-hoc effect sizes

```{r embryo-posthoc-eta}
eta2_from_groups <- function(a, b, label) {
  tt   <- t.test(a, b, var.equal = FALSE)
  eta2 <- tt$statistic^2 / (tt$statistic^2 + tt$parameter)
  cat(label, "— eta²:", round(eta2, 4), "\n")
}

wt_oil  <- df_ki67$Ki67_pct[df_ki67$Genotype == "Chd1+/+"     & df_ki67$Treatment == "Corn oil"]
wt_T    <- df_ki67$Ki67_pct[df_ki67$Genotype == "Chd1+/+"     & df_ki67$Treatment == "Testosterone"]
mut_oil <- df_ki67$Ki67_pct[df_ki67$Genotype == "Chd1R616Q/+" & df_ki67$Treatment == "Corn oil"]
mut_T   <- df_ki67$Ki67_pct[df_ki67$Genotype == "Chd1R616Q/+" & df_ki67$Treatment == "Testosterone"]

eta2_from_groups(mut_T,  mut_oil, "Mut testosterone vs Mut corn oil")
eta2_from_groups(mut_T,  wt_T,   "Mut testosterone vs WT testosterone")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
