---
title: "CHD1_AR_motif"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CHD1 CUT&RUN motif analysis
First, I selected the top 50,000 most significant CHD1 peaks (lowest p-values) from the WT analysis (called using MACS2 from combined bed files, using q < 0.01 for peak calling) and ran through AME in the online MEME suite tool with the following settings:

* Sequence motif score: avg_odds
* HOCOMOCO v12 known motifs (Human and mouse orthologs)
* Background model source: letter frequencies in (primary) sequences file (20240816_CHD1_WT_DMSO_merged_peaks_to_50k.bed.txt.fa)
* Background model frequencies: 0.28529,0.21471,0.21471,0.28529
* Total pseudocount added to a motif column: 0.1
* Statistical test: Fisher's exact test (optimized over motif scores)
* Labeling positives: all 10000 primary sequences are labeled as 'positive'; all 10000 control sequences are labeled as 'negative'
* Classifying positives: sequences with the motif scores ≥ 'TP Thresh' are classified as positive
* E-value threshold for reporting results: 10


```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Reading in files

Here we read in the outputs we got from AME in online MEME suite:

```{r reading_in_files}
# Path to your AME tsv file(s)
# If you have multiple groups, list them all with names
ame_files <- list(
  "MUT_DMSO" = "C:/Users/kimberl/OneDrive - Menntaský/HTB Lab/Bioinformatics analyses/20240410_Decode_cutandrun/Merged_CHD1_peaks/20240816_CHD1_R616Q_DMSO_merged_q01/ame (1).tsv",
  "MUT_DHT"  = "C:/Users/kimberl/OneDrive - Menntaský/HTB Lab/Bioinformatics analyses/20240410_Decode_cutandrun/Merged_CHD1_peaks/20240816_CHD1_R616Q_DHT_merged_q01/ame.tsv",
  "WT_DMSO" = "C:/Users/kimberl/OneDrive - Menntaský/HTB Lab/Bioinformatics analyses/20240410_Decode_cutandrun/Merged_CHD1_peaks/20240816_CHD1_WT_DMSO_merged_q01/ame_new_50k.tsv"
  )


# Output file
outfile <- "AME_bubble_plot.pdf"

##############################################################################
# LOAD AND PARSE AME FILES
##############################################################################

read_ame <- function(path, group_name) {
  # Skip comment lines starting with #
  lines <- readLines(path)
  data_lines <- lines[!grepl("^#", lines)]
  df <- read.table(
    text  = paste(data_lines, collapse = "\n"),
    header = TRUE,
    sep    = "\t",
    quote  = "",
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  df$group <- group_name
  return(df)
}

ame_list <- list()
for (name in names(ame_files)) {
  ame_list[[name]] <- read_ame(ame_files[[name]], name)
}
ame_data <- bind_rows(ame_list)

# Clean up column names (remove special characters)
colnames(ame_data) <- make.names(colnames(ame_data))
# Rename key columns for convenience
ame_data <- ame_data %>%
  rename(
    motif_ID    = motif_ID,
    pvalue      = p.value,
    adj_pvalue  = adj_p.value,
    evalue      = E.value
  )
```

## Filter to motifs of interest

```{r}
# Motifs to plot — use the motif_ID column values
# Edit this list to whatever you want to show
motifs_to_plot <- c(
  "ANDR.H12CORE.2.P.B",    # Androgen receptor — key for reviewer
  "FOXA1.H12CORE.0.P.B",   # FOXA1 pioneer factor
  "SP1.H12CORE.2.P.B",     # SP1
  "KLF4.H12CORE.0.PSM.A"    # KLF4
  # add or remove motifs here
)

plot_data <- ame_data %>%
  dplyr::filter(motif_ID %in% motifs_to_plot) %>%
  mutate(
    # Clean up motif names for display — take just the TF name before first dot
    motif_label = sub("\\..*", "", motif_ID),
    # Fix AR label
    motif_label = ifelse(motif_label == "ANDR", "AR", motif_label),
    # log10 transform for plotting
    log10_pvalue = -log10(pvalue),
    log10_evalue = -log10(evalue),
    # Factor for ordering — use the order in motifs_to_plot
    motif_label = factor(motif_label,
                         levels = rev(sub("\\..*", "",
                                          gsub("ANDR", "AR", motifs_to_plot)))),
    group = factor(group, levels = names(ame_files))
  )

# Warn if any requested motifs are missing
missing <- setdiff(motifs_to_plot, ame_data$motif_ID)
if (length(missing) > 0) {
  message("Warning: these motifs were not found in the AME output:\n",
          paste(missing, collapse = "\n"))
}
```

## Plot

```{r}
p <- ggplot(plot_data,
            aes(x     = group,
                y     = motif_label,
                size  = log10_evalue,
                color = log10_pvalue)) +
  geom_point(alpha = 0.9) +
  scale_size_continuous(
    name   = expression(-log[10](E-value)),
    range  = c(2, 12),
    breaks = pretty(plot_data$log10_evalue, n = 4)
  ) +
  scale_color_gradientn(
    name   = expression(-log[10](p-value)),
    colors = c("grey90", "#CBC3E3", "#9B59B6", "#6C3483", "#1A0533"),
    guide  = guide_colorbar(barwidth = 1, barheight = 6)
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y      = element_text(color = "black"),
    axis.title       = element_blank(),
    axis.line        = element_line(color = "grey40"),
    axis.ticks       = element_line(color = "grey40"),
    panel.background = element_rect(fill = "white"),
    plot.background  = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.text      = element_text(color = "black"),
    legend.title     = element_text(color = "black"),
    panel.grid.major = element_line(color = "grey", linewidth = 0.3),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Motif enrichment at CHD1 CUT&RUN peaks"
  ) +
  theme(plot.title = element_text(color = "white", hjust = 0.5))

print(p)

ggsave(outfile, p, width = 5, height = 4, device = "pdf")
message("Plot saved to: ", outfile)

# Also print a quick summary table of the plotted values
plot_data %>%
  dplyr::select(group, motif_label, pvalue, adj_pvalue, evalue) %>%
  arrange(group, pvalue) %>%
  as.data.frame()

plot_data

# Plot WT_DMSO only
plot_data_wt <- plot_data %>%
  dplyr::filter(group == "WT_DMSO")

plot_data_wt

p_wt <- ggplot(plot_data_wt,
               aes(x     = group,
                   y     = motif_label,
                   size  = log10_evalue,
                   color = log10_pvalue)) +
  geom_point(alpha = 0.9) +
  scale_size_continuous(
    name   = expression(-log[10](E-value)),
    range  = c(2, 12),
    limits = c(0, max(plot_data_wt$log10_pvalue)),
    breaks = c(0, pretty(plot_data_wt$log10_evalue, n = 3))
  ) +
  scale_color_gradient(
    low    = "grey",
    high   = "darkblue",
    name   = expression(-log[10](p-value)),
    limits = c(0, max(plot_data_wt$log10_pvalue)),
    guide  = guide_colorbar(barwidth = 1, barheight = 6)
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x       = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y       = element_text(color = "black"),
    axis.title        = element_blank(),
    axis.line         = element_line(color = "grey40"),
    axis.ticks        = element_line(color = "grey40"),
    panel.background  = element_rect(fill = "white"),
    plot.background   = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.text       = element_text(color = "black"),
    legend.title      = element_text(color = "black"),
    panel.grid.major  = element_line(color = "grey", linewidth = 0.3),
    panel.grid.minor  = element_blank(),
    plot.title        = element_text(color = "black", hjust = 0.5),
    plot.caption      = element_text(
      color = "black",
      hjust = 0,
      size = 10
    )
  ) +
  labs(
    title = "Motif enrichment at CHD1 CUT&RUN peaks",
    caption = "Figure legend: Motif enrichment among CHD1 CUT&RUN peaks. Each point represents an enriched motif within a peak group. Point size indicates enrichment strength as -log10(E-value), with larger points representing lower E-values and stronger motif enrichment. Point colour indicates statistical significance as -log10(p-value), with darker purple representing more significant enrichment."
  )
print(p_wt)

ggsave("AME_bubble_plot_WT_DMSO.pdf",
       p_wt, width = 3, height = 4, device = "pdf")


ame_data <- ame_data %>%
  mutate(
    frac_TP = TP / pos,
    frac_FP = FP / neg,
    fold_enrichment = frac_TP / frac_FP,
    odds_ratio = (TP * (neg - FP)) / (FP * (pos - TP))
  )
print(ame_data)
```

