---
title: "CHD1 KO RNA-seq analysis for PILBOS manuscript"
author: "Kimberley Jade Anderson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

# Overview

This R Markdown contains only the CHD1 KO NPC RNA-seq analyses used for the manuscript section on androgen rescue of CHD1-deficiency-associated transcriptional changes.

This script generates:

- DESeq2 results for:
  - CHD1 KO vs control, DMSO-treated NPCs
  - DHT vs DMSO in CHD1 KO NPCs
  - DHT + enzalutamide vs DHT in CHD1 KO NPCs
  - DHT + enzalutamide vs DMSO in CHD1 KO NPCs
- QC plots for the RNA-seq dataset
- Gene lists for downstream interpretation
- Log2FC correlation plots used for the DHT rescue and enzalutamide analyses
- Correlation comparison statistics testing whether enzalutamide attenuates DHT rescue

# Libraries

```{r libraries}
library(tidyverse)
library(GenomicFeatures)
library(tximport)
library(DESeq2)
library(EnsDb.Mmusculus.v79)
library(psych)
library(RColorBrewer)
library(pheatmap)
```

# Import kallisto abundance files

```{r import-kallisto, cache=TRUE}
# Build TxDb from the same GTF used for kallisto
txdb <- makeTxDbFromGFF(
  "Mus_musculus.GRCm39.115.gtf/Mus_musculus.GRCm39.115.gtf",
  format = "gtf"
)

tx_keys <- keys(txdb, keytype = "TXNAME")

tx2gene <- AnnotationDbi::select(
  txdb,
  keys = tx_keys,
  columns = "GENEID",
  keytype = "TXNAME"
)

samples <- dir(path = "kallisto_output/", pattern = "_Kallisto")

sample.sheet <- data.frame(samples = samples) %>%
  mutate(
    replicate = sub("^(A[0-9]+)_.*", "\\1", samples),
    genotype = sub("^A[0-9]+_([^_]+)_.*", "\\1", samples),
    treatment = sub("^A[0-9]+_[^_]+_([^_]+)_.*", "\\1", samples),
    treatment = ifelse(grepl("D_E", samples), "D_E", treatment),
    genotype = ifelse(grepl("CHD1", samples), "CHD1_KO", genotype),
    group = factor(paste(genotype, treatment, sep = "_")),
    replicate = factor(replicate)
  )

sample.sheet$group <- relevel(sample.sheet$group, ref = "CTL_DMSO")

files <- file.path("kallisto_output", samples, "abundance.h5")
names(files) <- sample.sheet$samples

txi <- tximport(
  files,
  type = "kallisto",
  tx2gene = tx2gene,
  ignoreTxVersion = TRUE
)
```

# DESeq2 analysis

```{r deseq2, cache=TRUE}
ddsTxi <- DESeqDataSetFromTximport(
  txi,
  colData = sample.sheet,
  design = ~ replicate + group
)

# Keep genes expressed with count >= 2 in at least 3 samples in at least one group
keep <- apply(counts(ddsTxi), 1, function(x) {
  any(tapply(x >= 2, ddsTxi$group, sum) >= 3)
})

dds <- ddsTxi[keep, ]
dds <- DESeq(dds)

resultsNames(dds)

# Main manuscript contrasts
res_genotype <- results(
  dds,
  contrast = c("group", "CHD1_KO_DMSO", "CTL_DMSO")
)

res_KO_DHT_vs_DMSO <- results(
  dds,
  contrast = c("group", "CHD1_KO_DHT", "CHD1_KO_DMSO")
)

res_KO_DEnz_vs_DHT <- results(
  dds,
  contrast = c("group", "CHD1_KO_D_E", "CHD1_KO_DHT")
)

res_KO_DEnz_vs_DMSO <- results(
  dds,
  contrast = c("group", "CHD1_KO_D_E", "CHD1_KO_DMSO")
)

normalized.counts <- counts(dds, normalized = TRUE)
colnames(normalized.counts) <- sample.sheet$samples
```

# RNA-seq QC

```{r qc-plots}
hist(
  res_genotype$pvalue,
  breaks = 100,
  main = "Histogram of p-values: CHD1 KO vs control",
  xlab = "p-value",
  freq = FALSE,
  col = "grey70"
)

hist(
  res_KO_DHT_vs_DMSO$pvalue,
  breaks = 100,
  main = "Histogram of p-values: DHT vs DMSO in CHD1 KO",
  xlab = "p-value",
  freq = FALSE,
  col = "grey70"
)

vsd <- vst(dds, blind = FALSE)

sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)

rownames(sampleDistMatrix) <- paste(vsd$group, vsd$replicate, sep = "_")
colnames(sampleDistMatrix) <- rownames(sampleDistMatrix)

heatmap.cols <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  col = heatmap.cols
)

plotPCA(vsd, intgroup = "group")
plotPCA(vsd, intgroup = "replicate")

plotMA(res_genotype, ylim = c(-6, 6), main = "CHD1 KO vs control")
plotMA(res_KO_DHT_vs_DMSO, ylim = c(-6, 6), main = "DHT vs DMSO in CHD1 KO")
```

# Add gene symbols and export results

```{r annotate-and-export-results}
gene.annotation <- ensembldb::select(
  EnsDb.Mmusculus.v79,
  keys = rownames(dds),
  keytype = "GENEID",
  columns = c("GENEID", "SYMBOL")
) %>%
  distinct(GENEID, .keep_all = TRUE)

add_gene_symbols <- function(res) {
  res.df <- as.data.frame(res) %>%
    rownames_to_column("ensembl_id") %>%
    mutate(
      gene = gene.annotation$SYMBOL[
        match(ensembl_id, gene.annotation$GENEID)
      ]
    ) %>%
    arrange(pvalue) %>%
    dplyr::filter(!is.na(padj))

  return(res.df)
}

res.genotype.df <- add_gene_symbols(res_genotype)
res.dht.df <- add_gene_symbols(res_KO_DHT_vs_DMSO)
res.dht.enz.df <- add_gene_symbols(res_KO_DEnz_vs_DHT)
res.dmso.dht.enz.df <- add_gene_symbols(res_KO_DEnz_vs_DMSO)

write.csv(
  res.genotype.df,
  "CHD1_KO_vs_control_DESeq2_results.csv",
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  res.dht.df,
  "CHD1_KO_DHT_vs_DMSO_DESeq2_results.csv",
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  res.dht.enz.df,
  "CHD1_KO_DHT_ENZ_vs_DHT_DESeq2_results.csv",
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  res.dmso.dht.enz.df,
  "CHD1_KO_DHT_ENZ_vs_DMSO_DESeq2_results.csv",
  row.names = FALSE,
  quote = FALSE
)

sig.genotype <- res.genotype.df %>% dplyr::filter(padj < 0.1)
sig.dht <- res.dht.df %>% dplyr::filter(padj < 0.1)
sig.dht.enz <- res.dht.enz.df %>% dplyr::filter(padj < 0.1)
sig.dmso.dht.enz <- res.dmso.dht.enz.df %>% dplyr::filter(padj < 0.1)
```

# Prepare p-value deciles

```{r pvalue-deciles}
res.genotype.df <- res.genotype.df %>%
  mutate(pval.decile = ntile(pvalue, 10))

res.dht.df <- res.dht.df %>%
  mutate(pval.decile = ntile(pvalue, 10))

res.dht.enz.df <- res.dht.enz.df %>%
  mutate(pval.decile = ntile(pvalue, 10))

res.dmso.dht.enz.df <- res.dmso.dht.enz.df %>%
  mutate(pval.decile = ntile(pvalue, 10))

topdecile.genotype <- res.genotype.df %>% dplyr::filter(pval.decile == 1)
topdecile.dht <- res.dht.df %>% dplyr::filter(pval.decile == 1)
topdecile.dht.enz <- res.dht.enz.df %>% dplyr::filter(pval.decile == 1)
topdecile.dmso.dht.enz <- res.dmso.dht.enz.df %>% dplyr::filter(pval.decile == 1)

write.csv(
  topdecile.genotype,
  "CHD1_KO_vs_control_top_pvalue_decile.csv",
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  topdecile.dht,
  "CHD1_KO_DHT_vs_DMSO_top_pvalue_decile.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# Plotting functions

```{r plotting-functions}
theme_chd1_paper <- function(base_size = 8) {
  theme_classic(base_size = base_size) +
    theme(
      text = element_text(family = "Arial", colour = "black"),
      axis.title = element_text(size = base_size, colour = "black"),
      axis.text = element_text(size = base_size - 1, colour = "black"),
      axis.line = element_line(linewidth = 0.4, colour = "black"),
      axis.ticks = element_line(linewidth = 0.4, colour = "black"),
      plot.title = element_text(size = base_size + 1, hjust = 0.5),
      panel.border = element_blank(),
      legend.position = "none"
    )
}

plot_lfc_correlation <- function(res.x,
                                 res.y,
                                 xlab,
                                 ylab,
                                 title,
                                 file,
                                 decile = 1,
                                 lims = NULL) {

  x.decile <- res.x %>% dplyr::filter(pval.decile == decile)
  y.decile <- res.y %>% dplyr::filter(pval.decile == decile)

  lfc.compare <- inner_join(
    x.decile,
    y.decile,
    by = "ensembl_id",
    suffix = c("_x", "_y")
  ) %>%
    dplyr::filter(!is.na(log2FoldChange_x), !is.na(log2FoldChange_y))

  cor.stats <- cor.test(
    lfc.compare$log2FoldChange_x,
    lfc.compare$log2FoldChange_y,
    method = "pearson"
  )

  r.value <- unname(cor.stats$estimate)
  p.value <- cor.stats$p.value

  if (is.null(lims)) {
    max.lim <- ceiling(max(abs(c(
      lfc.compare$log2FoldChange_x,
      lfc.compare$log2FoldChange_y
    )), na.rm = TRUE))
    lims <- c(-max.lim, max.lim)
  }

  p.label <- ifelse(
    p.value < 2.2e-16,
    "'< 2.2e-16'",
    paste0("'= ", signif(p.value, 2), "'")
  )

  p <- ggplot(
    lfc.compare,
    aes(x = log2FoldChange_x, y = log2FoldChange_y)
  ) +
    geom_point(
      shape = 21,
      size = 1.4,
      stroke = 0.25,
      colour = "black",
      fill = "black"
    ) +
    geom_vline(xintercept = 0, linewidth = 0.35) +
    geom_hline(yintercept = 0, linewidth = 0.35) +
    geom_smooth(
      method = lm,
      se = TRUE,
      colour = "#E69F00",
      fill = "grey75",
      linewidth = 0.7
    ) +
    coord_fixed(xlim = lims, ylim = lims) +
    theme_chd1_paper(base_size = 8) +
    ggtitle(paste0(title, " (n = ", nrow(lfc.compare), ")")) +
    xlab(xlab) +
    ylab(ylab) +
    annotate(
      "text",
      x = Inf,
      y = Inf,
      hjust = 1.1,
      vjust = 1.4,
      size = 3,
      label = paste0(
        "italic(r) == ", round(r.value, 3),
        "*','~~italic(p)~", p.label
      ),
      parse = TRUE
    )

  print(p)

  ggsave(
    file,
    p,
    width = 3.2,
    height = 2.6,
    units = "in",
    device = cairo_pdf
  )

  return(list(
    data = lfc.compare,
    plot = p,
    r = r.value,
    p = p.value,
    n = nrow(lfc.compare)
  ))
}
```

# Figure 4E: CHD1 KO transcriptional signature is reversed by DHT

```{r figure-4e-ko-dht-rescue}
fig4e.ko.dht <- plot_lfc_correlation(
  res.x = res.genotype.df,
  res.y = res.dht.df,
  xlab = expression(Log[2] * "FC CHD1 KO vs control"),
  ylab = expression(Log[2] * "FC DHT vs DMSO in CHD1 KO"),
  title = "CHD1 KO vs DHT response",
  file = "Figure_4E_CHD1_KO_vs_DHT_decile1.pdf",
  decile = 1,
  lims = c(-1.5, 1.5)
)

write.csv(
  fig4e.ko.dht$data,
  "Figure_4E_CHD1_KO_vs_DHT_decile1_genes.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# Supplementary analysis: DHT response opposed by enzalutamide

```{r supplementary-dht-vs-enz}
sfig.dht.enz <- plot_lfc_correlation(
  res.x = res.dht.df,
  res.y = res.dht.enz.df,
  xlab = expression(Log[2] * "FC DHT vs DMSO in CHD1 KO"),
  ylab = expression(Log[2] * "FC DHT + ENZ vs DHT in CHD1 KO"),
  title = "DHT response vs ENZ response",
  file = "Supp_Figure_DHT_vs_ENZ_decile1.pdf",
  decile = 1,
  lims = c(-1.5, 1.5)
)

write.csv(
  sfig.dht.enz$data,
  "Supp_Figure_DHT_vs_ENZ_decile1_genes.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# Supplementary analysis: DHT + ENZ partially attenuates rescue

```{r supplementary-ko-vs-dht-enz}
sfig.ko.dht.enz <- plot_lfc_correlation(
  res.x = res.genotype.df,
  res.y = res.dmso.dht.enz.df,
  xlab = expression(Log[2] * "FC CHD1 KO vs control"),
  ylab = expression(Log[2] * "FC DHT + ENZ vs DMSO in CHD1 KO"),
  title = "CHD1 KO vs DHT + ENZ response",
  file = "Supp_Figure_CHD1_KO_vs_DHT_ENZ_decile1.pdf",
  decile = 1,
  lims = c(-2.5, 2.5)
)

write.csv(
  sfig.ko.dht.enz$data,
  "Supp_Figure_CHD1_KO_vs_DHT_ENZ_decile1_genes.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# All-gene correlations and Steiger's test

```{r all-gene-correlations-and-steiger}
all.lfc <- list(
  genotype = res.genotype.df %>%
    dplyr::select(ensembl_id, genotype_lfc = log2FoldChange, pval.decile),
  dht = res.dht.df %>%
    dplyr::select(ensembl_id, dht_lfc = log2FoldChange),
  dht_enz = res.dmso.dht.enz.df %>%
    dplyr::select(ensembl_id, dht_enz_lfc = log2FoldChange)
) %>%
  purrr::reduce(inner_join, by = "ensembl_id") %>%
  dplyr::filter(
    !is.na(genotype_lfc),
    !is.na(dht_lfc),
    !is.na(dht_enz_lfc)
  )

all.lfc.decile1 <- all.lfc %>%
  dplyr::filter(pval.decile == 1)

all.gene.r.dht <- cor(
  all.lfc$genotype_lfc,
  all.lfc$dht_lfc,
  use = "complete.obs"
)

all.gene.r.dht.enz <- cor(
  all.lfc$genotype_lfc,
  all.lfc$dht_enz_lfc,
  use = "complete.obs"
)

decile1.r.dht <- cor(
  all.lfc.decile1$genotype_lfc,
  all.lfc.decile1$dht_lfc,
  use = "complete.obs"
)

decile1.r.dht.enz <- cor(
  all.lfc.decile1$genotype_lfc,
  all.lfc.decile1$dht_enz_lfc,
  use = "complete.obs"
)

# Steiger's test comparing dependent correlations:
# genotype vs DHT compared with genotype vs DHT + ENZ
steiger.all.genes <- psych::r.test(
  n = nrow(all.lfc),
  r12 = all.gene.r.dht,
  r13 = all.gene.r.dht.enz,
  r23 = cor(all.lfc$dht_lfc, all.lfc$dht_enz_lfc, use = "complete.obs")
)

steiger.decile1 <- psych::r.test(
  n = nrow(all.lfc.decile1),
  r12 = decile1.r.dht,
  r13 = decile1.r.dht.enz,
  r23 = cor(
    all.lfc.decile1$dht_lfc,
    all.lfc.decile1$dht_enz_lfc,
    use = "complete.obs"
  )
)

steiger.all.genes
steiger.decile1

correlation.summary <- tibble(
  analysis = c(
    "All genes: genotype vs DHT",
    "All genes: genotype vs DHT+ENZ",
    "Top genotype p-value decile: genotype vs DHT",
    "Top genotype p-value decile: genotype vs DHT+ENZ"
  ),
  n = c(
    nrow(all.lfc),
    nrow(all.lfc),
    nrow(all.lfc.decile1),
    nrow(all.lfc.decile1)
  ),
  r = c(
    all.gene.r.dht,
    all.gene.r.dht.enz,
    decile1.r.dht,
    decile1.r.dht.enz
  )
)

write.csv(
  correlation.summary,
  "CHD1_KO_DHT_ENZ_correlation_summary.csv",
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  all.lfc,
  "CHD1_KO_DHT_ENZ_all_gene_lfc_table.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# Session information

```{r session-info}
sessionInfo()
```
