---
title: "CHD1 CUT&RUN analysis for Chd1R616Q NPCs"
author: "Kimberley Jade Anderson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This R Markdown contains the CHD1 CUT&RUN analysis steps relevant to the manuscript figures and statements:

- Bowtie2 alignment-rate summary
- Peak-calling workflow documentation
- DiffBind setup and consensus CHD1 peakset generation
- Peak annotation
- Overlap between CHD1-bound genes and the top RNA-seq p-value decile
- Export of gene lists used for downstream interpretation

Exploratory analyses and unused sections have been removed.

# Libraries

```{r libraries}
library(tidyverse)
library(knitr)
library(kableExtra)
library(DiffBind)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.ensGene)
library(org.Mm.eg.db)
library(Mus.musculus)
library(VennDiagram)
library(GeneOverlap)

txdb <- TxDb.Mmusculus.UCSC.mm10.ensGene
```

# Alignment QC

The original FASTQ processing, Bowtie2 alignment, SAM/BAM conversion, bigWig generation, and MACS3 peak calling were performed outside R. The commands are retained here for reproducibility but are not evaluated during knitting.

```{bash bowtie2-fastqc, eval=FALSE}
# Run FastQC
find . -maxdepth 1 -mindepth 1 -type d -exec sh -c "cd '{}' && pwd && fastqc *.fastq.gz -t 90" \;

# Align paired-end reads to mm10 using Bowtie2
find . -maxdepth 1 -mindepth 1 -type d -exec sh -c "cd '{}' && \
  bowtie2 -p 40 --time --met-stderr --met-file bt2_metrics.txt \
  -x /hpcdata/Mimir/kimberl/mm10_bt2/mm10 \
  -1 '{$1}'_L008_R1_001.fastq.gz \
  -2 '{$1}'_L008_R2_001.fastq.gz \
  -S '{$1}'.sam" \;
```

```{bash sam-to-bam, eval=FALSE}
# Convert SAM files to coordinate-sorted BAM files
for i in `ls */*.sam | rev | cut -c 5- | rev | uniq`
do
  samtools view -@ 16 -bS ${i}.sam | \
  samtools sort -T ${i} -O BAM -@ 16 -o ${i}.bam
done
```

```{bash macs3-peak-calling, eval=FALSE}
# Merge IgG files for shared control
samtools merge -o all_merged_IgGs.bam -@ 48 *IgG*.bam

# Activate MACS3 environment
ml load Anaconda3/2022.05
conda activate /hpcapps/env/conda/MACS3

# Call CHD1 peaks against merged IgG control
for i in `ls *CHD1*.bam | rev | cut -c 5- | rev | uniq`
do
  echo "Calling peaks for sample: $i"
  macs3 callpeak \
    -t ${i}.bam \
    -c all_merged_IgGs.bam \
    -n ${i} \
    -f BAMPE \
    -g mm \
    --outdir ${i} \
    -q 0.1 \
    --max-gap 300 \
    --keep-dup all
done
```

## Bowtie2 alignment-rate table

```{r alignment-rate-table, eval=file.exists("20240411_bt2_output.txt"), cache=TRUE}
bt2_file <- "20240411_bt2_output.txt"

bt2_lines <- readLines(bt2_file)

alignment.table <- tibble(line = bt2_lines) %>%
  mutate(
    is_sample = str_detect(line, "^/"),
    sample = if_else(is_sample, basename(line), NA_character_)
  ) %>%
  tidyr::fill(sample) %>%
  filter(str_detect(sample, regex("CHD1|IgG", ignore_case = TRUE))) %>%
  filter(str_detect(line, "overall alignment rate")) %>%
  mutate(
    alignment_rate = str_extract(
      line,
      "\\d+\\.\\d+(?=% overall alignment rate)"
    ) %>% as.numeric()
  ) %>%
  dplyr::select(sample, alignment_rate) %>%
  arrange(sample)

alignment.table %>%
  kable(
    caption = "Bowtie2 overall alignment rates for CHD1 and IgG CUT&RUN samples",
    col.names = c("Sample", "Overall alignment rate (%)"),
    digits = 2
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )

write.csv(
  alignment.table,
  "CHD1_CUTRUN_bowtie2_alignment_rates.csv",
  row.names = FALSE
)
```

# BigWig files and TSS profiles

These commands generate normalized bigWig files and CHD1 signal profiles over transcription start sites. They are retained as documentation of the figure-generation workflow.

```{bash bigwig-and-tss-profile, eval=FALSE}
module load deepTools
module load SAMtools

samtools index -M *merged.bam

# Generate RPGC-normalized bigWig files
for i in `ls *merged.bam | rev | cut -c 5- | rev | uniq`
do
  bamCoverage \
    --bam ${i}.bam \
    -o ${i}_depthnorm.bw \
    -p max \
    --binSize 5 \
    --smoothLength 30 \
    --normalizeUsing RPGC \
    --effectiveGenomeSize 2652783500 \
    --ignoreForNormalization chrX \
    --minFragmentLength 130 \
    --extendReads \
    --centerReads
done

# Signal over TSSs
computeMatrix reference-point \
  --referencePoint TSS \
  -o CHD1_overTSSs \
  -S 20240816_CHD1_WT_DMSO_merged_depthnorm.bw \
     20240816_CHD1_R616Q_DMSO_merged_depthnorm.bw \
     20240816_CHD1_R616Q_DHT_merged_depthnorm.bw \
  -R mm10_transcripts.bed \
  -a 3000 \
  -b 3000 \
  -p 24 \
  --samplesLabel WT Mut_DMSO Mut_DHT

plotProfile -m CHD1_overTSSs -o CHD1_overTSSs.pdf
plotProfile -m CHD1_overTSSs -o CHD1_overTSSs_olap.pdf --plotType overlapped_lines
```

# DiffBind sample sheet

```{r sample-sheet, eval=dir.exists("Peak_calling") && file.exists("Genotypes.csv"), cache=TRUE}
samples <- list.dirs("Peak_calling/", full.names = FALSE, recursive = FALSE)

diffbind.samplesheet <- data.frame(
  SampleID = samples[grepl("CHD1", samples)]
)

diffbind.samplesheet <- diffbind.samplesheet %>%
  tidyr::separate_wider_delim(
    SampleID,
    delim = "_",
    names = c("Embryo", "Treatment", "Factor", "Sample"),
    cols_remove = FALSE
  )

genotypes <- read.csv2("Genotypes.csv")

diffbind.samplesheet <- dplyr::left_join(
  diffbind.samplesheet,
  genotypes,
  by = NULL
)

# Check what the genotype column is called
names(diffbind.samplesheet)

# Rename columns safely
if ("Genotype" %in% names(diffbind.samplesheet)) {
  diffbind.samplesheet <- diffbind.samplesheet %>%
    dplyr::rename(Condition = Genotype)
}

if ("Embryo" %in% names(diffbind.samplesheet)) {
  diffbind.samplesheet <- diffbind.samplesheet %>%
    dplyr::rename(Tissue = Embryo)
}

if ("Sample" %in% names(diffbind.samplesheet)) {
  diffbind.samplesheet <- diffbind.samplesheet %>%
    dplyr::select(-Sample)
}

diffbind.samplesheet <- diffbind.samplesheet %>%
  dplyr::mutate(
    bamReads = paste0("bam_files/", SampleID, ".bam"),
    controlID = "all_merged_IgGs",
    bamControl = "bam_files/all_merged_IgGs.bam",
    Peaks = paste0("Peak_calling/", SampleID, "/", SampleID, "_peaks.xls"),
    PeakCaller = "macs"
  )

write.csv(
  diffbind.samplesheet,
  "diffBind_sampleSheet_generated.csv",
  row.names = FALSE
)

diffbind.samplesheet
```

# DiffBind analysis

This section creates a DiffBind object, counts reads in peaks, applies blacklist/greylist filtering, normalizes, tests for differential CHD1 occupancy, and exports the WT DMSO consensus peakset.

```{r diffbind-analysis, eval=file.exists("diffBind_sampleSheet_1.csv"), cache=TRUE}
final.samplesheet <- read.csv2("diffBind_sampleSheet_1.csv") %>%
  filter(str_detect(SampleID, "CHD1"))

dba.all <- dba(sampleSheet = final.samplesheet)
plot(dba.all)

dba.all <- dba.count(dba.all)
```

```{r, cache=TRUE}
# Remove blacklist regions and generate/apply greylist
dba.all <- dba.blacklist(dba.all, greylist = FALSE)

# Normalize using native DiffBind methods with background signal
chd1.norm.background <- dba.normalize(
  dba.all,
  method = DBA_ALL_METHODS,
  normalize = DBA_NORM_NATIVE,
  background = TRUE
)

# Test for differential occupancy by condition
chd1.dba <- dba.analyze(
  chd1.norm.background,
  method = DBA_ALL_METHODS,
  design = "~Condition"
)

dba.show(chd1.dba, bContrasts = TRUE)

# Diagnostic plots
dba.plotMA(chd1.dba, contrast = 1, method = DBA_DESEQ2)
dba.plotMA(chd1.dba, contrast = 1, method = DBA_EDGER, bNormalized = TRUE, th = 0.1)
dba.plotVolcano(chd1.dba, contrast = 1, method = DBA_EDGER, th = 0.1)

# Export all tested sites
chd1.diff.peaks <- dba.report(chd1.dba, th = 1, method = DBA_EDGER)
chd1.diff.peaks.df <- as.data.frame(chd1.diff.peaks)

write.csv(
  chd1.diff.peaks.df,
  "CHD1_diffPeaks_all_tested_sites.csv",
  row.names = FALSE
)

# Retrieve WT DMSO consensus peaks
chd1.wt.peaks <- dba.peakset(
  chd1.dba,
  chd1.dba$masks$`Chd1+/+` & chd1.dba$masks$DMSO,
  bRetrieve = TRUE,
  DataType = DBA_DATA_FRAME
)

write.table(
  chd1.wt.peaks[, 1:3],
  "CHD1_WT_DMSO_consensus_peaks.bed",
  col.names = FALSE,
  row.names = FALSE,
  sep = "\t",
  quote = FALSE
)
```

# Merged WT CHD1 peakset

For the manuscript, the CHD1-bound gene list was generated from merged WT DMSO CHD1 peaks, filtered by MACS score.

```{r merged-wt-peakset, eval=file.exists("Merged_CHD1_peaks/20240816_CHD1_WT_DMSO_merged_q01/20240816_CHD1_WT_DMSO_merged_peaks.narrowPeak"), cache=TRUE}
merged.wt.peaks <- read.table(
  "Merged_CHD1_peaks/20240816_CHD1_WT_DMSO_merged_q01/20240816_CHD1_WT_DMSO_merged_peaks.narrowPeak",
  header = FALSE
)

# MACS narrowPeak column 5 is the score.
merged.wt.peaks.filtered <- merged.wt.peaks %>%
  filter(V5 > 100)

write.table(
  merged.wt.peaks.filtered[, 1:4],
  "merged_WT_CHD1_peaks_filtered.bed",
  col.names = FALSE,
  row.names = FALSE,
  quote = FALSE,
  sep = "\t"
)
```

# Annotate CHD1 peaks

```{r score-distribution, cache=FALSE}
raw <- read.table(
  "Merged_CHD1_peaks/20240816_CHD1_WT_DMSO_merged_q01/20240816_CHD1_WT_DMSO_merged_peaks.narrowPeak",
  header = FALSE
)

cat("Score quantiles:\n")
print(quantile(raw$V5, probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1)))
cat("\nPeaks with V5 > 200:", sum(raw$V5 > 200), "\n")
cat("\nPeaks with V5 > 100:", sum(raw$V5 > 100), "\n")
cat("Peaks with V5 > 50:", sum(raw$V5 > 50), "\n")
cat("Peaks with V5 > 10:", sum(raw$V5 > 10), "\n")
cat("Total peaks:", nrow(raw), "\n")
```

This will show us whether `V5 > 200` is genuinely filtering out a meaningful proportion of peaks or whether almost all peaks already exceed that threshold.

```{r annotate-peaks}
anno_function <- function(bedfile) {
  cuttag_anno <- annotatePeak(
    bedfile,
    TxDb = txdb,
    annoDb = "Mus.musculus"
  )
  plotAnnoPie(cuttag_anno)
  as.data.frame(cuttag_anno)
}

CHD1_WT_peaks_anno <- anno_function("Merged_CHD1_peaks/20240816_CHD1_WT_DMSO_merged_q01/20240816_CHD1_WT_DMSO_merged_peaks.narrowPeak")
nrow(CHD1_WT_peaks_anno)

#Limit CHD1 peaks to only with score over 200
CHD1_WT_peaks_anno <- CHD1_WT_peaks_anno[CHD1_WT_peaks_anno$V5 >200,]
nrow(CHD1_WT_peaks_anno)
write.table(CHD1_WT_peaks_anno[,c(1:4)], "merged_WT_CHD1_peaks_filtered.bed", 
                col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")
CHD1_WT_peaks_anno <- anno_function("merged_WT_CHD1_peaks_filtered.bed")

cat("Peaks after filter:", nrow(CHD1_WT_peaks_anno), "\n")
cat("Unique genes:", n_distinct(CHD1_WT_peaks_anno$SYMBOL), "\n")


# DEGs
genotype_DEGs <- read.csv("20231116_Chd1_NPC_genotype_DEGs.csv")
genotype_DEGs$pval.decile <- ntile(genotype_DEGs$cor.pval, 10)
genotype_DEGs_top.dec <- genotype_DEGs %>% filter(pval.decile == 1)

# Unique gene sets
chd1.bound.genes <- CHD1_WT_peaks_anno %>%
  filter(!is.na(SYMBOL), SYMBOL != "") %>%
  distinct(SYMBOL)

deg.top.genes <- genotype_DEGs_top.dec %>%
  filter(!is.na(gene), gene != "") %>%
  distinct(gene)

overlap.genes <- intersect(CHD1_WT_peaks_anno$SYMBOL, deg.top.genes$gene)

cat("CHD1-bound genes:", nrow(chd1.bound.genes), "\n")
cat("Top decile DEGs:", nrow(deg.top.genes), "\n")
cat("Overlap:", length(overlap.genes), "\n")
```

# Overlap between CHD1-bound genes and RNA-seq results

This analysis asks how many genes in the top p-value decile from the genotype RNA-seq comparison are directly associated with CHD1 CUT&RUN peaks.

```{r cutrun-rnaseq-overlap}


overlap.genes <- intersect(chd1.bound.genes$SYMBOL, deg.top.genes$gene)

cat("CHD1-bound genes:", nrow(chd1.bound.genes), "\n")
cat("Top decile DEGs:", nrow(deg.top.genes), "\n")
cat("Overlap:", length(overlap.genes), "\n")
```

## Venn diagram

```{r cutrun-rnaseq-venn}

chd1.rnaseq.venn <- draw.pairwise.venn(
  area1      = nrow(chd1.bound.genes),
  area2      = nrow(deg.top.genes),
  cross.area = length(overlap.genes),
  euler.d    = TRUE,
  scaled     = TRUE,
  ext.text   = TRUE,
  category   = c("CHD1-bound genes", "Top RNA-seq p-value decile")
)

pdf("CHD1_CUTRUN_RNAseq_top10percent_venn.pdf", height = 1.5, width = 1.5, pointsize = 6)
grid.draw(chd1.rnaseq.venn)
dev.off()
```

## Fisher's exact test for overlap

```{r cutrun-rnaseq-fisher,cache=TRUE}
gene.overlap <- newGeneOverlap(
  chd1.bound.genes$SYMBOL,
  deg.top.genes$gene,
  spec = "mm9.gene"
)
gene.overlap <- testGeneOverlap(gene.overlap)
gene.overlap
```

# Session information

```{r session-info}
sessionInfo()
```
